---
title: "Introduction to CamelRatiosIndex"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to CamelRatiosIndex}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  message = FALSE,
  comment = "#>",
  fig.width = 9,
  fig.height = 6
)
```

## Overview

The **CamelRatiosIndex** package implements the multivariate-weighted indexing method proposed by Ayimah et al. (2023a, 2023b) for bank performance assessment using the CAMEL framework. The package provides:

- **`camel_index()`**: Computes composite year-on-year indices from CAMEL ratio data
- **`plot_camel_index()`**: Visualizes percentage differences across banks using ggplot2
- **Built-in data**: Example datasets from Ghanaian commercial banks (2015-2022)

This composite index is intended to offer regulators and policymakers a standardised, objective for monitoring bank performance over time and across institutions. Its ability to benchmark banks against a common base year enhances early-warning capabilities, enabling supervisory authorities to identify emerging weaknesses individual banks as well as systemic vulnerabilities within the industry.

## The CAMEL Framework

CAMEL is an internationally recognized framework for evaluating bank performance, comprising five dimensions:

| Dimension | Ratio | Direction |
|-----------|-------|-----------|
| **C**apital Adequacy | Ca | Higher = better |
| **A**sset Quality | Aq | Higher = worse (inverted) |
| **M**anagement Efficiency | Me | Higher = worse (inverted) |
| **E**arnings | Eq | Higher = better |
| **L**iquidity | Lm | Higher = worse (inverted) |

## Installation

```r
# Install official release from CRAN
install.packages("CamelRatiosIndex")

# Install from GitHub (development version)
# install.packages("remotes")
remotes::install_github("JC-Ayimah/CamelRatiosIndex")
```

## Quick Start

### Computing the CAMEL Index

```{r}
library(CamelRatiosIndex)

# Load built-in example data
data("camel_2015")
data("camel_2022")

# Compute the index
result <- camel_index(camel_2015, camel_2022)

# View the main output
result$index_table
```

### Accessing Detailed Results

```{r}
# Laspeyres-type indices (base year weights)
result$mw_lasp

# Paasche-type indices (current year weights)
result$mw_pash

# Communality weights from base year factor analysis
result$weights_base

# Eigenvalues
result$eigenvalues_base
```

### Visualizing Results

```{r}
# Basic plot
plot_camel_index(result)

# Highlight specific banks
plot_camel_index(result, highlight_banks = c("Absa", "Ecobank", "GCB"))

# Custom styling
plot_camel_index(
  result,
  title = "Ghanaian Banks: 2015 vs 2022",
  subtitle = "CAMEL Index Performance",
  theme_fn = ggplot2::theme_bw
)
```

## Data Format

### Data Frame Input

When using data frames, the first column must be the bank identifier, followed by the five CAMEL ratios:

```{r}
# Example structure
head(camel_2015)
```

### Matrix Input

For matrices, supply bank names separately:

```{r}
base_mat <- as.matrix(camel_2015[, -1])
curr_mat <- as.matrix(camel_2022[, -1])
banks <- camel_2015$Bank

result2 <- camel_index(base_mat, curr_mat, bank_names = banks)
```

## Understanding the Output

The `camel_index()` function returns a rich object with multiple components:

```{r}
# Print overview
print(result)

# Detailed summary
summary(result)
```

### Key Metrics

- **I_mw**: Composite index (base = 100). Values > 100 indicate improvement; < 100 indicate decline.
- **PD**: Percentage difference from base year. Positive = improvement.
- **mw_lasp**: Laspeyres-type index using base year communality weights.
- **mw_pash**: Paasche-type index using current year communality weights.
- **weights_base/current**: Communality values from robust factor analysis, representing the proportion of variance explained by each CAMEL ratio.

## Trend Analysis: Year-on-Year Comparison

The package supports trend analysis by computing indices for successive pairs of years. This allows you to track bank performance trajectories over multiple periods and identify emerging patterns that a single comparison might miss.

### Computing Year-on-Year Indices

The package includes built-in datasets for eight years (2015-2022). To build a trend, compute the index for each consecutive year pair, using the earlier year as the base:

```{r}
# Compute year-on-year indices
yoy_results <- list(
  "2015-2016" = camel_index(camel_2015, camel_2016, n_factors = 2),
  "2016-2017" = camel_index(camel_2016, camel_2017, n_factors = 2),
  "2017-2018" = camel_index(camel_2017, camel_2018, n_factors = 2),
  "2018-2019" = camel_index(camel_2018, camel_2019, n_factors = 2),
  "2019-2020" = camel_index(camel_2019, camel_2020, n_factors = 2),
  "2020-2021" = camel_index(camel_2020, camel_2021),
  "2021-2022" = camel_index(camel_2021, camel_2022, n_factors = 2)
)
```

### Extracting Trend Data

Extract the index values for each bank across all periods:

```{r}
library(dplyr)
library(purrr)

# Build a trend table: one row per bank, one column per year-pair
trend_data <- yoy_results |>
  imap(\(data, period) {
    data$index_table |>
      select(bank, !!period := I_mw)
  }) |>
  reduce(full_join, by = "bank")

# View the trend table
trend_data
```

### Visualizing Trends

Plot the trajectory of a specific bank over time:

```{r}
library(ggplot2)
library(tidyr)

# Reshape for ggplot2
trend_long <- trend_data |>
  pivot_longer(
    cols = -bank,
    names_to = "period",
    values_to = "I_mw"
  ) |>
  mutate(
    end_year = as.numeric(gsub(".*-", "", period))
  )

# Plot trends for selected banks
selected_banks <- c("Absa", "Ecobank", "GCB", "ADB")

trend_long |>
  filter(bank %in% selected_banks) |>
  ggplot(aes(x = end_year, y = I_mw, colour = bank, group = bank)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2) +
  geom_hline(yintercept = 100, linetype = "dashed", color = "grey50") +
  scale_x_continuous(breaks = 2016:2022) +
  labs(
    title = "CAMEL Index Trends: Selected Ghanaian Banks",
    subtitle = "Year-on-year composite indices (base = 100 for each period)",
    x = "Year",
    y = "Composite Index (I_mw)",
    colour = "Bank"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    plot.title = element_text(face = "bold", size = 14)
  )
```

### Industry-Wide Trend

Compute the mean index across all banks for each period to assess overall sector health:

```{r}
# Industry mean index by period
industry_trend <- trend_long |>
  group_by(end_year) |>
  summarise(
    mean_I_mw = mean(I_mw, na.rm = TRUE),
    sd_I_mw = sd(I_mw, na.rm = TRUE),
    n_banks = n(),
    .groups = "drop"
  )

industry_trend

# Plot industry trend with confidence band
industry_trend |>
  ggplot(aes(x = end_year, y = mean_I_mw)) +
  geom_ribbon(
    aes(ymin = mean_I_mw - sd_I_mw, ymax = mean_I_mw + sd_I_mw),
    fill = "steelblue", alpha = 0.2
  ) +
  geom_line(colour = "steelblue", linewidth = 1) +
  geom_point(colour = "steelblue", size = 2.5) +
  geom_hline(yintercept = 100, linetype = "dashed", colour = "grey50") +
  scale_x_continuous(breaks = 2016:2022) +
  labs(
    title = "Industry-Wide CAMEL Index Trend",
    subtitle = "Mean composite index across all banks ± 1 SD",
    x = "Year",
    y = "Mean Composite Index (I_mw)"
  ) +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold", size = 14))
```

### Identifying Consistent Performers

Find banks that improved consistently (index > 100 in most periods):

```{r}
# Count how many periods each bank improved
trend_summary <- trend_long |>
  group_by(bank) |>
  summarise(
    periods_above_100 = sum(I_mw > 100, na.rm = TRUE),
    total_periods = sum(!is.na(I_mw)),
    mean_I_mw = mean(I_mw, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    improvement_rate = round(periods_above_100 / total_periods * 100, 1)
  ) |>
  arrange(desc(improvement_rate), desc(mean_I_mw))

trend_summary
```


## Methodology

The index computation follows these steps:

1. **Transform ratios**: Invert Aq, Me, and Lm so that higher values always indicate better performance.
2. **Compute correlations**: Build correlation matrices for base and current years.
3. **Extract eigenvalues**: Determine the number of factors (eigenvalues > 1).
4. **Robust factor analysis**: Use [robustfa::FaCov()] with OGK covariance estimation.
5. **Extract communalities**: Use communalities as weights for each CAMEL ratio.
6. **Compute indices**: Calculate Laspeyres and Paasche indices, then average them.
7. **Scale to base 100**: Final composite index with percentage difference.

## References

Ayimah, J. C., Mettle, F. O., Nortey, E. N., & Minkah, R. (2023a). A Robust Multivariate Weighting Technique for Computing a Measure for Inflation. *African Journal of Technical Education and Management*, 3(1), 1-15. Retrieved from	https://ajtem.com/index.php/ajtem/article/view/53. 

Ayimah, J.C. (2023b). Computing Multivariate-Weighted Consumer Price Index: An Application Manual in R. B P International. DOI: 10.9734/bpi/mono/978-81-19315-32-1.
DOI:http://dx.doi.org/10.9734/bpi/mono/978-81-19315-32-1
